---
title: "Installing bonsaiR: macOS and Linux"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing bonsaiR: macOS and Linux}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

`bonsaiR` wraps a real, non-trivial systems stack -- conda, git, a
compiled C++ binary (Sanity), a compiled Cython/C extension (cellstates),
and MPI -- not just an R package with CRAN dependencies. This vignette is
purely about getting that stack set up correctly; see
`vignette("bonsaiR-intro")` for actually using the package once installed.

**Platform note up front:** the macOS instructions below were worked out
by actually hitting and fixing every failure mode described, on real
Apple Silicon hardware. The Linux instructions are based on
`bonsai_install()`'s own coded logic plus one real debugging session on an
institutional HPC cluster (the shared-conda-cache issue below) -- solid,
but less exhaustively battle-tested than the macOS path. If you hit
something not covered here on Linux, it's genuinely useful information;
consider it worth reporting.

The two platforms differ in exactly one important way: **macOS's system
compiler has no OpenMP support at all**, which Linux's does by default.
Most of the macOS-specific steps below exist solely to work around that;
skip straight to the Linux section if that's your platform.

## Common prerequisites (both platforms)

- **R** (4.x), with `devtools` installed (`install.packages("devtools")`)
- **conda or Miniconda**, with the `conda` command on your shell's `PATH`
- **git**

## macOS installation

### 1. Xcode Command Line Tools

Needed for `git` and basic build tools. If you don't already have them:

```{bash, eval = FALSE}
xcode-select --install
```

### 2. Homebrew

If you don't already have it: <https://brew.sh>

### 3. A real C/C++ compiler with OpenMP support

**This is the step that matters most on macOS.** Apple's system `clang`
(the default `cc`/`gcc` you get without doing anything) has *no* OpenMP
support whatsoever -- not a missing flag, not a config issue, it simply
cannot compile `-fopenmp` code. Both Sanity's C++ build and cellstates'
Cython extension need real OpenMP. Install Homebrew's GCC instead:

```{bash, eval = FALSE}
brew install gcc
```

You don't need to configure anything further yourself --
`bonsai_install()` detects Homebrew GCC automatically (by scanning for a
versioned `gcc-N`/`g++-N` pair under `/opt/homebrew/opt/gcc/bin` or
`/usr/local/opt/gcc/bin`) and uses it specifically for the steps that need
OpenMP.

### 4. System OpenMPI

Needed if you want to run the core tree search with `use_mpi = TRUE`
(recommended above ~2000 cells):

```{bash, eval = FALSE}
brew install openmpi
```

### 5. A known gotcha: a broken `mpicc` can shadow the working one

If you also have Anaconda (not just Miniconda) installed, its base
environment can ship its own `mpicc` wrapper script that expects a
conda-toolchain compiler (e.g. `arm64-apple-darwin20.0.0-clang`) which was
never actually installed -- and depending on your shell's `PATH` order,
*that* broken `mpicc` can be found before the working Homebrew one,
causing cryptic build failures. This is a real bug that was hit and fixed
during this package's own development, on this exact kind of setup.

Check which `mpicc` your shell finds first, and whether it actually works:

```{bash, eval = FALSE}
command -v mpicc
```

```{bash, eval = FALSE}
mpicc --version
```

If that fails with something like `arm64-apple-darwin20.0.0-clang:
command not found`, you've hit this issue. `bonsai_install()` itself
works around it internally (it explicitly forces the known-good compiler
to the front of `PATH` for the specific build steps that need it), so
**this doesn't block `bonsai_install()` from working** -- but if you're
also pre-installing packages manually (see "Pre-installing the conda
environment" below), you need to apply the same fix yourself, since a
plain `pip install mpi4py` will just use whatever `mpicc` is first on
`PATH`.

### 6. Install and set up bonsaiR

```{r, eval = FALSE}
devtools::install_github("gevensen95/bonsaiR")
library(bonsaiR)
benv <- bonsai_install()
```

This clones and compiles everything (slow the first time, fast on every
subsequent call since each step is skipped if it's already done).

### 7. Verify

```{r, eval = FALSE}
bonsai_smoke_test(bonsai_env = benv)
```

## Linux installation

### 1. Compiler and MPI packages

Unlike macOS, Linux's system `gcc` supports OpenMP natively -- there is no
Homebrew-GCC-equivalent workaround needed here. On Debian/Ubuntu:

```{bash, eval = FALSE}
sudo apt-get install build-essential libgomp1 libopenmpi-dev openmpi-bin git
```

`libgomp1` is the OpenMP runtime specifically; `bonsai_install()` checks
for it before attempting to build Sanity and gives a clear error naming
this exact package if it's missing.

On other distributions, install the equivalent of: a C/C++ compiler
(`gcc`/`g++`), the OpenMP runtime, an OpenMPI development package
(providing `mpicc`), and `git`.

### 2. conda/Miniconda

If you don't already have conda, install Miniconda:
<https://docs.conda.io/projects/miniconda/en/latest/>

### 3. A known gotcha: shared/HPC conda installs and permission errors

If you're on an institutional HPC cluster with a shared, multi-user conda
installation (common -- look for a `conda info` output showing a `base
environment` path owned by IT/a shared group rather than you personally),
you may hit a permission error like this when `bonsai_install()` (or a
manual `conda create`) tries to write to the shared package cache:

```
PermissionError: [Errno 13] Permission denied:
  '/path/to/shared/miniconda3/pkgs/cache/....info.json'
```

This is not a bonsaiR bug -- it means conda is trying to write a lock/state
file into a shared cache directory you don't have write access to. Fix it
by pointing conda at directories you actually own (this only edits your
personal `~/.condarc`, not the shared system config, so it needs no admin
permissions and doesn't affect other users of the shared install):

```{bash, eval = FALSE}
conda config --add pkgs_dirs ~/.conda/pkgs
```

```{bash, eval = FALSE}
conda config --add envs_dirs ~/.conda/envs
```

Then retry. This fixes both a manual `conda create` and `bonsai_install()`
itself, since both resolve through the same `pkgs_dirs`/`envs_dirs` search
order.

### 4. Install and set up bonsaiR

```{r, eval = FALSE}
devtools::install_github("gevensen95/bonsaiR")
library(bonsaiR)
benv <- bonsai_install()
```

### 5. Verify

```{r, eval = FALSE}
bonsai_smoke_test(bonsai_env = benv)
```

## Pre-installing the conda environment ahead of time

On either platform, you can set up just the conda environment and Python
packages in advance (e.g. to save time later, or on a machine with
limited internet access) without running the full `bonsai_install()`.
This uses the exact same environment name, Python version, and pinned
package versions `bonsai_install()` itself expects, so a later
`bonsai_install()` call detects everything as already done and skips
straight to cloning and compiling Sanity/cellstates -- the one part of
setup that has to happen fresh on every machine regardless, since it
produces compiled, architecture-specific output.

```{bash, eval = FALSE}
conda create -n bonsai python=3.11 -y
source "$(conda info --base)/etc/profile.d/conda.sh"
conda activate bonsai
pip install --upgrade matplotlib==3.11.0 numpy==2.4.6 scipy==1.17.1 pandas==3.0.3 \
  scikit-learn==1.9.0 ruamel.yaml==0.19.1 psutil==7.2.2 shiny==1.6.3 faicons==0.2.2 \
  shinyswatch==0.11.0 jinja2==3.1.6 tables==3.11.1 h5py==3.16.0 natsort==8.4.0
export MPICC="$(command -v mpicc)"
export PATH="$(dirname "$MPICC"):$PATH"
pip install --upgrade mpi4py==4.0.0
```

(This is the same sequence walked through step-by-step earlier in this
package's development, packaged into one block here -- on macOS,
double-check `$MPICC` is the *working* `mpicc` first, per the gotcha
above, since a plain `command -v mpicc` isn't guaranteed to find it.)

## Caching the environment object

`bonsai_install()`'s return value (`benv`) is a plain list of file paths
with nothing live or unserializable in it -- safe to cache so you don't
need to re-run the (fast, but not free) existence checks every session:

```{r, eval = FALSE}
benv_path <- "~/.bonsaiR_env.rds"
if (file.exists(benv_path)) {
  benv <- readRDS(benv_path)
} else {
  benv <- bonsai_install()
  saveRDS(benv, benv_path)
}
```

This is only valid on the machine it was created on, and goes stale if you
later run `bonsai_install(force = TRUE)` or otherwise change the
underlying environment -- if functions using a cached `benv` start failing
with path-not-found errors, regenerate it first.

## Troubleshooting checklist

If `bonsai_install()` fails partway through, it's almost always one of:

- **No `mpicc` on `PATH` at all** -- install system OpenMPI (`brew install
  openmpi` / `apt-get install openmpi-bin libopenmpi-dev`).
- **An `mpicc` is found, but it's broken** -- macOS-specific; see the
  gotcha above. Check with `mpicc --version`, not just `command -v mpicc`.
- **`libgomp1`/OpenMP runtime missing** -- Linux-specific; the error
  message from `bonsai_install()` names this package directly.
- **Permission errors writing to a shared conda install** -- HPC-specific;
  see the gotcha above.
- **A stale conda environment from a previous failed attempt** -- since
  every step is idempotent, first just re-run `bonsai_install(force =
  FALSE)` (the default) to retry only the step that failed. Only use
  `force = TRUE` (which rebuilds everything from scratch) if that doesn't
  work.

For anything else, `bonsai_install()`'s own output is streamed live and
usually names the actual failing command -- the fastest path to a fix is
usually reading the last ~20 lines of that output rather than guessing.
