---
title: "Getting Started with bonsaiR: A Real-Data Walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bonsaiR: A Real-Data Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette walks through the full `bonsaiR` pipeline on real (if small)
single-cell data: [Seurat](https://satijalab.org/seurat/)'s built-in
`pbmc_small` object, an 80-cell, 230-gene subsample of a real 10x Genomics
PBMC dataset. It ships with Seurat itself, so this vignette needs no extra
downloads.

The pipeline takes a `Seurat` object with raw UMI counts and produces a
maximum-likelihood tree representation of the cells (Bonsai, Moriel et al.
2025), via:

1. **Sanity** -- Bayesian normalization that estimates, for every gene in
   every cell, both a point estimate of true expression and an error bar on
   that estimate.
2. **cellstates** -- pre-clustering cells into statistically
   indistinguishable groups, used to "warm-start" the tree search.
3. **Bonsai** -- the core maximum-likelihood tree reconstruction.

Each step is a separate external tool (two Python, one C++), which is why
`bonsai_install()` exists: it builds a dedicated conda environment and
compiles the C++/Cython pieces for you, once.

`pbmc_small` is realistic in shape (real 10x PBMC counts, real dropout,
real gene-expression noise) but still small enough that this entire
vignette -- including the actual Bonsai tree search -- runs in well under
half an hour on a laptop. **This is not representative of runtime at
Bonsai's intended scale.** Bonsai's own documentation targets tens of
thousands of cells, at which point the core tree search should be run with
MPI across many cores (or on an HPC cluster), not interactively. This
vignette deliberately uses `use_mpi = FALSE`: at 80 cells there is nothing
to gain from parallelizing, and single-core avoids [a known intermittent
race condition in Bonsai's own core-calculation code](../reference/run_bonsai.html)
that has only been observed on small MPI runs like this one -- see
`?run_bonsai` for details.

## One-time setup

`bonsai_install()` creates a conda environment (`bonsai` by default) and
clones + builds the Sanity and cellstates source repositories. This step
is slow the first time (several minutes, more if a compiler toolchain
needs installing) and instant on every subsequent call, since it skips any
step that's already done.

```{r setup}
library(bonsaiR)
library(Seurat)

benv <- bonsai_install()
benv
```

## The data

```{r load-data}
data("pbmc_small")
pbmc_small

# pbmc_small ships with pre-computed cluster identities from Seurat's own
# tutorial -- we'll reuse these later to define marker-gene comparison
# groups and to color the tree.
table(Idents(pbmc_small))
```

## Step 1: Sanity normalization

`bonsai_write_sanity_input()` extracts raw counts from the Seurat object
and writes them in the sparse Matrix Market format Sanity expects (with
rows sorted, which Sanity's own parser requires but `Matrix::writeMM()`
does not guarantee on its own). `run_sanity()` then normalizes them,
producing per-gene, per-cell expression estimates with error bars.

```{r sanity}
work_dir <- tempfile("bonsaiR_pbmc_small_")
dir.create(work_dir)

sanity_in <- bonsai_write_sanity_input(
  pbmc_small,
  output_dir = file.path(work_dir, "sanity_input")
)
sanity_in

sanity_out <- run_sanity(
  sanity_in, benv,
  output_dir = file.path(work_dir, "sanity_output")
)
sanity_out
```

## Step 2: cellstates pre-clustering

cellstates operates on the same raw counts (not Sanity's normalized
output) to find cells that are statistically indistinguishable given
measurement noise alone. `bonsai_write_config()` uses this clustering to
build a "warm-start" tree, which substantially speeds up Bonsai's search
compared to starting from a plain star tree.

```{r cellstates}
cellstates_out <- run_cellstates(
  sanity_in, benv,
  output_dir = file.path(work_dir, "cellstates_output")
)
cellstates_out
```

## Step 3: configure and run Bonsai

```{r config}
config <- bonsai_write_config(
  sanity_output = sanity_out,
  cellstates_output = cellstates_out,
  bonsai_env = benv,
  dataset_name = "pbmc_small",
  results_folder = file.path(work_dir, "bonsai_results")
)
config
```

```{r run-bonsai}
result <- run_bonsai(config, benv, use_mpi = FALSE)
result
```

## Step 4: read the reconstructed tree

```{r read-tree}
tree <- bonsai_read_tree(result, benv)
tree

ape::Ntip(tree$phylo)   # number of leaves (cells)
ape::Nnode(tree$phylo)  # number of internal nodes
```

A quick look at the tree, colored by the original Seurat clusters:

```{r plot-tree, fig.width=7, fig.height=7}
cluster_by_cell <- setNames(as.character(Idents(pbmc_small)), colnames(pbmc_small))
tip_clusters <- cluster_by_cell[tree$phylo$tip.label]
tip_colors <- c("0" = "#4477AA", "1" = "#EE6677", "2" = "#228833")[tip_clusters]

ape::plot.phylo(tree$phylo, type = "fan", show.tip.label = FALSE)
ape::tiplabels(pch = 16, col = tip_colors, cex = 0.6)
legend("topleft", legend = names(table(tip_clusters)), pch = 16,
       col = tip_colors[!duplicated(tip_clusters)], title = "Seurat cluster")
```

If Bonsai's tree reconstruction agrees with Seurat's own clustering, cells
from the same cluster should mostly fall into the same clade(s) rather
than being scattered across the tree.

## Step 5: marker genes between two clades

`bonsai_marker_genes()` computes, for every gene, the probability that a
randomly picked cell from group 1 has higher expression than a randomly
picked cell from group 2 -- Bonsai's marker score. Scores near 0 or 1
indicate strong markers. Here we compare Seurat clusters `0` and `2`.

Note this runs `bonsai_scout_preprocess.py` automatically the first time
it's called for a given results folder (building the visualization data
`calc_marker_genes.py` needs, which `run_bonsai()` does not produce on its
own) -- this is skipped on subsequent calls.

`bonsai_marker_genes()` also merges cells with (near-)zero branch-length
distance into single composite vertices before comparing groups (the same
merging the Bonsai-scout visualization app does), and will error out if
your two groups happen to split one of those merged clusters, rather than
silently comparing something misleading -- see `?bonsai_marker_genes` for
details. Clusters `0` and `2` here don't hit this; clusters `0` and `1`
do, since a couple of `pbmc_small`'s cells landed in different Seurat
clusters despite Bonsai finding them statistically indistinguishable.

```{r marker-genes}
group1_cells <- names(cluster_by_cell)[cluster_by_cell == "0"]
group2_cells <- names(cluster_by_cell)[cluster_by_cell == "2"]

markers <- bonsai_marker_genes(
  bonsai_tree = tree,
  group1_cells = group1_cells,
  group2_cells = group2_cells,
  bonsai_env = benv,
  bonsai_result = result,
  output_dir = file.path(work_dir, "marker_genes")
)

head(markers, 10)
```

The top hits are HLA class II genes (`HLA-DPB1`, `HLA-DPA1`, `HLA-DRB1`,
`HLA-DQA1`, ...) -- the textbook markers separating antigen-presenting
cells from other PBMC subsets, which is a good sign the reconstruction
recovered real biological structure rather than noise.

## Step 6: attach everything back to the Seurat object

For downstream analysis in your normal Seurat workflow, `bonsai_to_seurat()`
attaches the tree, posterior expression estimates, and cellstate labels
directly onto the Seurat object.

```{r to-seurat}
pbmc_small <- bonsai_to_seurat(
  pbmc_small, tree,
  cellstates_output = cellstates_out
)

# The full tree, plus per-cell posterior LTQs reordered to match the
# Seurat object's own cell order:
str(pbmc_small@misc$bonsai, max.level = 1)
dim(pbmc_small@misc$bonsai$leaf_posterior_ltqs)

# Cellstates cluster labels, as a regular metadata column:
table(pbmc_small$bonsai_cellstate)
```

## Next steps

- **Scaling up:** at tens of thousands of cells, re-run `run_bonsai()`
  with `use_mpi = TRUE` and a realistic `n_cores`, ideally on an HPC node
  -- see `?run_bonsai` and Bonsai's own README for guidance on scheduling
  this as a batch job rather than interactively.
- **A full smoke test:** `bonsai_smoke_test()` runs this entire pipeline
  end-to-end on synthetic data with built-in pass/fail reporting per
  stage -- useful for verifying a new `bonsai_install()` environment
  before pointing it at real data.
- **License note:** the underlying Bonsai tool stack is distributed under
  CC-BY-NC-4.0 -- not free for commercial use. Check this applies to your
  use case before relying on it.
