---
title: "Downstream Analyses with bonsaiR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Downstream Analyses with bonsaiR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette covers three functions for analyzing a Bonsai tree once
you've reconstructed it, building on the same `pbmc3k` example used in
`vignette("bonsaiR-intro")`:

- `run_average_over_groups()` -- noise-aware average expression per group,
  plus a ranking of which genes vary most across groups.
- `bonsai_cluster_by_annotation()` -- cuts the tree into clusters that best
  match a known annotation, and separately reports how well the tree's
  geometry alone (with no annotation at all) recovers those same labels.
- `bonsai_cluster_tree()` -- cuts the tree into clusters using geometry
  alone, with no annotation involved -- roughly `stats::cutree()`, but
  respecting the tree's actual branch lengths rather than just topology.

If you've already worked through `vignette("bonsaiR-intro")` you'll have
everything these functions need in your session already (`sanity_out`,
`cellstates_out`, `result`, `tree`, `pbmc3k`); this vignette re-derives
them from scratch so it also runs on its own.

## Setup

Same quick pipeline as the intro vignette -- see it for a full explanation
of each step.

```{r setup}
library(bonsaiR)
library(Seurat)

benv_path <- "~/.bonsaiR_env.rds"

if (file.exists(benv_path)) {
  benv <- readRDS(benv_path)
} else {
  benv <- bonsai_install()
  saveRDS(benv, benv_path)
}
work_dir <- tempfile("bonsaiR_downstream_")
dir.create(work_dir)

data("pbmc3k")
pbmc3k <- UpdateSeuratObject(pbmc3k)
Idents(pbmc3k) <- 'seurat_annotations'
cluster_by_cell <- setNames(as.character(Idents(pbmc3k)), colnames(pbmc3k))

sanity_in <- bonsai_write_sanity_input(pbmc3k, output_dir = file.path(work_dir, "sanity_input"))
sanity_out <- run_sanity(sanity_in, benv, output_dir = file.path(work_dir, "sanity_output"))
cellstates_out <- run_cellstates(sanity_in, benv, output_dir = file.path(work_dir, "cellstates_output"))

config <- bonsai_write_config(
  sanity_output = sanity_out,
  cellstates_output = cellstates_out,
  bonsai_env = benv,
  dataset_name = "pbmc3k",
  results_folder = file.path(work_dir, "bonsai_results")
)
result <- run_bonsai(config, benv, use_mpi = FALSE)
tree <- bonsai_read_tree(result, benv)
```

## Averaging expression over groups

`run_average_over_groups()` is the multi-group counterpart to
`bonsai_marker_genes()`'s pairwise two-group comparison -- instead of
testing one pair of clades, it summarizes every gene across *all* your
groups at once, weighting each cell by its own Sanity error bar (not a
plain mean).

Here we group by `pbmc3k`'s own Seurat clusters, though any named
grouping works the same way -- a Zone/cell-type annotation, cellstates
labels, or clades you've identified by eye in the tree plot.

```{r average-over-groups}
avg_result <- run_average_over_groups(
  sanity_out, cluster_by_cell, benv,
  output_dir = file.path(work_dir, "avg_by_cluster")
)
avg_result

# groups x genes: average expression per cluster
avg_result$avg_activities[, 1:5]

# per-gene ranking of how much expression varies across clusters
head(avg_result$significance, 10)
```

The top of `significance` should look familiar if you ran the marker-genes
step in the intro vignette: genes like `HLA-DPB1`, `HLA-DRB1`, `CST3`, and
`LYZ` tend to come out on top -- the same antigen-presenting/myeloid
markers, just found by summarizing across all clusters at once rather than
comparing one pair.

## Clustering the tree to match a known annotation

`bonsai_cluster_by_annotation()` cuts the tree to maximize Normalized
Mutual Information (NMI) against a given annotation. It also computes a
second clustering using *only* tree geometry (minimizing within-cluster
distance, never seeing your annotation at all) for comparison.

```{r cluster-by-annotation}
nmi_result <- bonsai_cluster_by_annotation(
  result, cluster_by_cell, benv,
  output_dir = file.path(work_dir, "nmi_clustering")
)
nmi_result

nmi_result$nmi_scores
head(nmi_result$clustering_results)
```

Read `nmi_scores` like this: the annotation-based row's NMI just confirms
the optimizer worked -- it was handed your labels and told to match them.
The **distance-based row's NMI is the informative number**: that
clustering never saw `cluster_by_cell` at all, so a high score there is a
genuine, quantitative measure of how well the tree's structure alone
recovers your known groups -- not something you can get from eyeballing a
tree plot.

## Clustering the tree by geometry alone

`bonsai_cluster_tree()` is the same distance-based clustering, exposed
directly rather than as a side-by-side comparison -- useful when you don't
have an annotation to compare against yet, or want to explore clusterings
at several granularities at once.

```{r cluster-tree}
tree_clusters <- bonsai_cluster_tree(result, n_clusters = 10, bonsai_env = benv)
str(tree_clusters)

# cluster sizes at the finest level computed
table(tree_clusters[[ncol(tree_clusters)]])
```

`tree_clusters` has one column per cut level, from a few clusters up to
`n_clusters` -- a byproduct of the underlying greedy-splitting algorithm.
Pick whichever column's granularity is useful for your analysis (e.g. via
`table()` on each column to see how cluster sizes change across levels),
the same way you'd choose a `k` when calling `stats::cutree()`.

## Collapsing and plotting a reduced tree

A tree plot at full single-cell scale is mostly noise -- thousands of
individual tips are impossible to read by eye. `bonsai_collapse_tree()`
reduces the tree to one representative tip per cluster (via
`ape::drop.tip()`, so branch lengths between cluster representatives stay
phylogenetically meaningful, not just a random subsample), and
`bonsai_plot_collapsed_tree()` plots the result with each point sized by
how many original cells it represents.

```{r collapse-tree, fig.width=6, fig.height=6}
finest <- tree_clusters[[ncol(tree_clusters)]]
tip_groups <- setNames(finest, rownames(tree_clusters))

reduced <- bonsai_collapse_tree(tree, tip_groups)
reduced

bonsai_plot_collapsed_tree(reduced)
```

By default each cluster is labeled with its raw geometry-cluster ID (e.g.
`cl_0`), which carries no biological meaning on its own -- it's just
whichever subtree the distance-based cut happened to produce. If you have
a separate annotation for the same cells, like `cluster_by_cell` from
above, `bonsai_dominant_labels()` finds each cluster's majority annotation
and `bonsai_rename_clusters()` applies it, so the plot shows what each
cluster actually *is* instead of an arbitrary ID:

```{r dominant-labels, fig.width=6, fig.height=6}
dominant <- bonsai_dominant_labels(reduced, cluster_by_cell)
dominant

reduced <- bonsai_rename_clusters(reduced, dominant)
bonsai_plot_collapsed_tree(reduced)
```

Each label's percentage is the dominant annotation's share of that
cluster's cells -- a cluster labeled e.g. `"CD14+ Mono (95%)"` is a clean,
well-separated group, while one at `"CD14+ Mono (55%)"` is a much more
mixed bag that's worth a closer look (e.g. at a finer-grained column of
`tree_clusters`) before trusting the label. If two different clusters end
up with the same dominant annotation, `bonsai_dominant_labels()`
disambiguates them automatically (appending the original cluster ID)
rather than producing a duplicate label `bonsai_rename_clusters()` would
otherwise reject.
